### 2018-03-27 submission/manuscript.Rmd
update discussion transitions

The majority of HiFi DNA polymerases we studied add a small increase in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some HiFi DNA polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community. Although avoiding HiFi DNA polymerases that yield high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.

Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach. However, bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen [@Meisel2016] but also can artifically increase the number of OTUs observed. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

>>>>

The majority of HiFi DNA polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some HiFi DNA polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community. Although avoiding HiFi DNA polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.

Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations. Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016]. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases may work better in certain situations over others. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

### 2018-03-27 submission/manuscript.tex
update discussion transitions

In this study we show that the number of OTUs, error rate, and chimera
prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
These differences are important because many diversity metrics rely on
the number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Importantly, many of these differences are due to undetected chimeras
that cannot be fully removed using standard bioinformatic approaches.
This suggests that some of the diversity differences between studies can
be attributed to differences in HiFi DNA polyermase used. Interestingly,
the earlier detection of differences in total number of OTUs between
HiFi DNA polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet (MSDS) the
differences between the two HiFi DNA polymerases are not immediately
apparent. Both HiFi DNA polymerases contain a recombinant \emph{Taq} DNA
polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. With everything else being equal, it is possible
that differences in how the recombinant \emph{Taq} was generated could
be a contributing factor for the differences observed between the HiFi
DNA polymerases.

There were few differences in bacterial community composition based on
HiFi DNA polymerase. The data also suggests that there were no
differences in the overall bacterial community composition for
sub-sampling depth or cycle number used. One possible reason for this
outcome was that our study did not have enough power to detect
differences due to low sample number in each group. Another reason was
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed HiFi DNA
polymerase dependent differences in chimera prevalence. Metrics that
simply look at presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas,
1996)) may be less robust to chimera prevalence and by extension total
number of OTU differences in HiFi DNA Polymerases. When choosing a
distance metric careful consideration of the biases introduced from the
PCR step of the 16S rRNA gene sequencing workflow need to be taken into
account.

Similar to using different preservation methods or different DNA
extraction kits, the type of HiFi DNA polymerase can add bias to the
observed bacterial community. The sequence error introduced by the HiFi
DNA polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. Within the larger context of the different 16S
rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used.

Heavy standardization has been commonly suggested as a reasonable answer
to finding the most reproducible approach. However, bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow. This study shows that specific diversity metrics
used to measure the microbial community consistently vary based on HiFi
DNA polymerase. Standardizing multiple workflows to one specific HiFi
DNA polymerase could be detrimental since the PCR step not only misses
entire species based on primer chosen (Meisel et al., 2016) but also can
artifically increase the number of OTUs observed. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.

>>>>

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on HiFi DNA polymerase {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between HiFi DNA polymerases in the mock versus fecal samples might
indicate that high biomass samples may underestimate the biases present
within low biomass samples. We observed that undetected chimeras that
were not identified and removed using standard bioinformatic approaches
cause many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in HiFi DNA
polyermase used. Based on our observations metrics that measure within
sample diversity depend on HiFi DNA polymerase but this may not be the
case for metrics that assess between sample diversity.

There were few differences that depend on HiFi DNA polymerase for
between sample diversity, as measured by the Bray-Curtis index. Using
this metric our observations found no differences in the overall
bacterial community composition for sub-sampling depth or cycle number
used. One possible reason for this outcome was that our study did not
have enough power to detect differences due to low sample number in each
group. Another reason was that many of the OTUs are likely not highly
abundant, allowing the Bray-Curtis index to be able to successfully
down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
diversity metric could be an important consideration in helping to
mitigate these observed HiFi DNA polymerase dependent differences in
chimera prevalence. Metrics that solely use presence/absence of OTUs
(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
prevalence and by extension total number of OTU differences in HiFi DNA
Polymerases. When choosing a distance metric careful consideration of
the biases introduced from the PCR step of the 16S rRNA gene sequencing
workflow need to be taken into account. With differences in the number
of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet (MSDS) the
differences between the two HiFi DNA polymerases are not immediately
apparent. Both HiFi DNA polymerases contain a recombinant \emph{Taq} DNA
polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. With everything else being equal, it is possible
that differences in how the recombinant \emph{Taq} was generated could
be a contributing factor for the differences observed between the HiFi
DNA polymerases. We are unlikely to avoid adding HiFi dependent bias to
16S rRNA gene sequencing results, however, these differences may not be
large enough to mask the actual biological signal.

The majority of HiFi DNA polymerases we studied add small increases in
the number of OTUs and chimera prevalence that can be masked by
biological differences. The sequence error introduced by the HiFi DNA
polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. The choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used because similar to using different preservation methods or
different DNA extraction kits, the type of HiFi DNA polymerase can add
bias to the observed bacterial community. Although avoiding HiFi DNA
polymerases that yield a high number of OTUs and chimera prevalence
might provide better standardization of results it does come at a cost.

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on HiFi DNA polymerase. Standardizing multiple workflows to one specific
HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
may work better in certain situations over others. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.
